Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-14 manuscript.tex
Final edits before sending to Andrei

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature. The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions. The magnitude of this displacement did not affect the results so long as all native WC bonds were completely broken. Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest. Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

@@

To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined purely as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
	\begin{equation}\label{e1}
	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow}
	\end{equation}
where $[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a^\mathrm{slow}$ is the association rate constant. In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d$ to solve for $k_d$ and $k_a$.
	\begin{equation}\label{e2a}
	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d(T_f)}{K_d(T_f)+4[S]_{T_f}}
	\end{equation}
	\begin{equation}\label{e2b}
	k_a^\mathrm{slow} = \frac{k_d^\mathrm{slow}}{K_d(T_f)}
	\end{equation}

FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement. 
	\begin{equation}\label{e3a}
	[S]_{T_f} = [1 - \theta(T_f)]c_{tot}
	\end{equation}
	\begin{equation}\label{e3b}
	K_d(T_f) = \frac{2c_{tot}[1 - \theta(T_f)]^2}{\theta(T_f)}
	\end{equation}
$c_{tot}$ is the total oligonucleotide concentration, which was 2 mM for all measurements.

@@

We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments. We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response. The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned Watson-Crick bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm. This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state. We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the Watson-Crick (WC) interaction. We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.

@@

We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models. MSMs are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates. We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences. Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium. By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F). In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates. The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model. The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways. Immediately we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior. In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM. These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates. The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization. Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization. The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways. The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes. The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization. We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.

@@

AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c). Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization. Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H. Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. Once in one of the four shifted states, the 5$^\prime$ vs. 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate. Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states. The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D. Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicC, Dickman2012ThermodynamicDNAs, DiMichele2014EffectHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D. The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state. A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states. Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time. A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s. As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.

@@

Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25-times more likely than dissociation F4 $\rightarrow$ D. Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation. The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability. We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation. Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a). Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s, which are again approximately equal.

@@

The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM. The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states. We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM. This stands in contrast to the metastable F4 state populated by GC-core, and can be understood by the reduced entropic stabilization offered by the frayed two-base AT tails relative to the frayed four-base AT-tails. Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s, which are again nearly equal.

>>>>

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997FastDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the nearest neighbor (NN) model predicted melting temperature.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions. The magnitude of this displacement did not affect the results so long as all native Watson-Crick (WC) bonds were completely broken. Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest. Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. In brief, a MSM extracts from simulation trajectories an ensemble of long-lived metastable macrostates, their equilibrium occupancy populations, and the equilibrium probability fluxes between them. In this manner, they provide an interpretable and predictive model of the system thermodynamics and long-time kinetics. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

@@

To determine observed rates from the T-jump data, the time-domain t-HDVE data was inverse Laplace transformed into the rate domain using a maximum entropy approach (MEM-iLT).~\citep{Kumar2001OnTimescales} Observed rates ($\lambda^\mathrm{fast}$ and $\lambda^\mathrm{slow}$) were computed from the amplitude-weighted mean rate across detected IR frequency, as previously described.~\citep{Sanstead2018DirectDehybridization} The fast response $k_d^\mathrm{fast}$ is defined as this amplitude-weighted mean rate, whereas the dissociation rate constant $k_d^\mathrm{slow}$ was extracted from the observed rate of the slow response $\lambda^\mathrm{slow}$ using a two-state model for self-complementary oligomers,~\citep{Bernasconi2012RelaxationKinetics}
	\begin{equation}\label{e1}
	\lambda^\mathrm{slow} = k_d^\mathrm{slow} + 4[S]_{T_f}k_a^\mathrm{slow},
	\end{equation}
where $[S]_{T_f}$ is the concentration of single-strand oligomer at the final temperature of the T-jump, and $k_a^\mathrm{slow}$ is the association rate constant. In practice, Eqn.~\ref{e1} is recast in terms of the dissociation equilibrium constant $K_d$ to solve for $k_d^\mathrm{slow}$ and $k_a^\mathrm{slow}$ as a function of $K_d$ and $[S]_{T_f}$,
	\begin{equation}\label{e2a}
	k_d^\mathrm{slow} = \frac{\lambda^\mathrm{slow}K_d(T_f)}{K_d(T_f)+4[S]_{T_f}},
	\end{equation}
	\begin{equation}\label{e2b}
	k_a^\mathrm{slow} = \frac{k_d^\mathrm{slow}}{K_d(T_f)}.
	\end{equation}

FTIR temperature series were measured for each sequence as reported previously,~\citep{Sanstead2016} and the second SVD component along temperature was fit to a two-state all-or-nothing model to determine the fraction of intact duplex ($\theta$) as a function of temperature.~\citep{Marky1987CalculatingCurves} $K_d$ and $[S]_{T_f}$ were then determined from $\theta$ at the final temperature of each T-jump measurement,
	\begin{equation}\label{e3a}
	[S]_{T_f} = [1 - \theta(T_f)]c_{tot},
	\end{equation}
	\begin{equation}\label{e3b}
	K_d(T_f) = \frac{2c_{tot}[1 - \theta(T_f)]^2}{\theta(T_f)},
	\end{equation}
where $c_{tot}$ is the total oligonucleotide concentration, which was 2 mM for all measurements.

@@

We first sought to demonstrate that the coarse-grained 3SPN.2 model accurately recapitulates the experimentally observed kinetics of DNA oligomer hybridization/dissociation by validating our computational predictions against temperature-jump infrared (T-jump IR) experiments. We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted the ``slow'' $k_d^\mathrm{slow}$ (10-30 $\mu$s) and ``fast'' $k_d^\mathrm{fast}$ (70-100 ns) rates estimated from the T-jump response. The slow response has previously been attributed to duplex dissociation on microsecond time scales induced by the rapid heating of the initially hybridized duplex.~\citep{Sanstead2016, Sanstead2018DirectDehybridization} The fast response has been assigned to terminal base pair fraying,~\citep{Sanstead2016, Sanstead2018DirectDehybridization} with this process corresponding to a relatively complex dynamical process that can span time scales from picoseconds to microseconds.~\cite{Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.~\citep{Pinamonti2019TheModels, Zgarbova2014BaseRNA}

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and fast $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm. This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state. We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction. We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.

@@

We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models. MSMs define the long-lived metastable macrostates of the system, their equilibrium occupancies, and the equilibrium transition probabilities between them. As such, they are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates. We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences. Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium. By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F). In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates. The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model. The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hydbrization/dehybridization pathways. Immediately, we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas at the other end of the spectrum GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior. In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM. These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates. The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization. Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization. The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways. The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes. The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization. We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.

@@

AT-all possesses the richest and most complex MSM of the four sequences by virtue of its repetitive nature, comprising a hybridized state (H), dissociated state (D), and four out-of-register shifted states (5S2, 3S2, 5S4, 3S4) (Fig.~\ref{fig:allseq_table}c). Analysis of the MSM transition probabilities reveal a critical role of the shifted states in mediating hybridization and dehybridization. Commencing from the dissociated state D, we observe approximately equal probabilities for transitions to each of the other five states, such that a transition to one of the out-of-register shifted states 5S2, 3S2, 5S4, or 3S4 is approximately 2.2 times more likely than a direct transition to the hybridized state H. Commencing from the hybridized state H, however, a direct transition to the dissociated state is approximately 1.2 times more likely than a transition to one of the two-base shifted states 5S2 or 3S2. Once in one of the four shifted states, the 5$^\prime$ vs.\ 3$^\prime$ overhang and degree of shifting play an important role in determining whether the duplex will transition to more shifted states, more aligned states, or completely dissociate. Transitions from more shifted states towards more aligned states (i.e., 5S4 $\rightarrow$ 5S2, 5S2 $\rightarrow$ H, 3S4 $\rightarrow$ 3S2, 3S2 $\rightarrow$ H) are approximately an order of magnitude more probable than the reverse transitions from more aligned states to more shifted states. The largest single transition probability from the four shifted states 5S2, 3S2, 5S4, and 3S4 is, however, back to the dissociated state D. Consistent with the higher destabilizing effect of 3$^\prime$ inert tails relative to 5$^\prime$ tails,~\citep{Doktycz1990ThermodynamicC, Dickman2012ThermodynamicDNAs, DiMichele2014EffectHybridization} the 3S4 $\rightarrow$ D transition probability is twice as large as the 5S4 $\rightarrow$ D, and the 3S2 $\rightarrow$ D is four times larger than the 5S2 $\rightarrow$ D. The transition probability from the 5S2 and 3S2 states back to the dissociated state D is equal to or greater than the transition probability to the hybridized state. A transition path theory analysis of the MSM reveals that 33\% of productive hybridization trajectories D $\dashrightarrow$ H (where the dashed arrow indicates the combination of both direct and indirect pathways) proceed through one or more out-of-register shifted states. Among these out-of-register pathways, the D $\dashrightarrow$ 5S2 $\dashrightarrow$ H transition is predicted to occur 57\% of the time. A mean first passage time (MFPT) analysis returns a MFPT for D $\dashrightarrow$ H of 3.0 $\mu$s and for H $\dashrightarrow$ D of 2.5 $\mu$s. As expected by the fact that the calculations are performed at the approximate melting temperature of the sequence, the MFPTs are approximately equal to one another.

@@

Analysis of the transition probabilities show that commencing from the F4 state, progression to the hybridized state F4 $\rightarrow$ H is 25 times more likely than dissociation F4 $\rightarrow$ D. Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed; concomitantly, H $\rightarrow$ F4 events tend to fall back to the H state and are unlikely to proceed to complete dissociation. The transition probabilities H $\rightarrow$ F4 and D $\rightarrow$ F4 are, respectively, one half and one quarter as likely as the F4 $\rightarrow$ H transition probability. We noted in Section \ref{sec:time} that fraying dynamics in the 3SPN.2 model appear to be significantly accelerated relative to center-of-mass translation, and it is conceivable that this may lead to elevated sampling of the F4 state within the computational model relative to experiment and the induction of more frequent dissociation. Moreover, since GC-core is the sequence most prone to fraying, this effect could be the root of the relatively poorer agreement of the $k_d^\mathrm{slow}$ response for GC-core compared to the other sequences due to an artificially elevated computational prediction of this rate (Fig.~\ref{fig:relaxation-comparison}a). Nevertheless, under these caveats our model predicts a MFPT for D $\rightarrow$ F4 $\rightarrow$ H of 3.4 $\mu$s and for H $\rightarrow$ F4 $\rightarrow$ D of 2.9 $\mu$s, which are again approximately equal.

@@

The GC-mix sequence is the only one of the four sequences studied that exhibits simple two-state ``all-or-nothing'' behavior.~\citep{Xiao2019, Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates, Sanstead2016} The GC-mix MSM comprises just two states, the hybridized H and dehybridized D (Fig.~\ref{fig:allseq_table}c), indicating that association and disassociation of the strands proceeds directly without passing through any metastable intermediate states resolvable under the $\tau$=1.2 ns lag time of our MSM. The two-state behavior appears to arise as a consequence of the placement of the G:C pair that maximally disrupts the repetitive AT tract within the decamer and destabilizing either out-of-register shifted states or frayed states. We note we do observe transient fraying of the terminal two-base AT tails within our dynamical simulations, but these frayed states are not sufficiently thermodynamically stable to produce a metastable macrostate within the resulting MSM. This stands in contrast to the metastable F4 state populated by GC-core. Our MSM predicts a MFPT for D $\rightarrow$ H of 2.9 $\mu$s and for H $\rightarrow$ D of 2.3 $\mu$s, which are again nearly equal.
